A Shared Subspace Model for Multi-Group Covariance Estimation

### 2019-02-22 paper/arxiv/mscov.tex
Membership predictions

\documentclass[12pt]{article}
\pdfoutput=1
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{latexsym}
\usepackage{textcomp}
\usepackage{longtable}
\usepackage{multirow, booktabs}
\usepackage{natbib}
\usepackage{url}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{chemarrow}
\usepackage{subcaption}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{tikz}
\usepackage[margin=.75in]{geometry}

@@

\title{Shared Subspace Models for Multi-Group Covariance Estimation
%
  \protect\thanks{Alexander M. Franks is a Moore/Sloan Data Science
    and WRF Innovation in Data Science Postdoctoral Fellow
    (\href{mailto:amfranks@uw.edu}{amfranks@uw.edu}).  Peter D. Hoff
    is a Professor in the Departments of Statistics and Biostatistics
    at the University of Washington
    (\href{mailto:pdhoff@uw.edu}{pdfhoff@uw.edu}).
%
This work was partially supported 
 by the Washington Research Foundation Fund for Innovation in
 Data-Intensive Discovery, the Moore/Sloan Data Science Environments
 Project at the University of Washington and NSF grant DMS-1505136.
%
 The authors are grateful to Dr.\ Daniel Promislow (Department of
 Pathology, University of Washington), Dr.\ Jessica Hoffman
 (University of Alabama at Birmingham) and Dr. Julie Josse (INRIA)
 for sharing data and ideas which contributed to framing of this
 paper.}}  \author{Alexander Franks and Peter Hoff} \date{\today}
\maketitle 
\begin{abstract}

% From Hoff Although the covariance matrices corresponding to different
% populations are unlikely to be exactly equal they can still exhibit a
% high degree of similarity. For example, some pairs of variables may be
% positively correlated across most groups, whereas the correlation
% between other pairs may be consistently negative. In such cases much
% of the similarity across covariance matrices can be described by
% similarities in their principal axes, which are the axes that are
% defined by the eigenvectors of the covariance matrices. Estimating the
% degree of across-popu- lation eigenvector heterogeneity can be helpful
% for a variety of estimation tasks.

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/3dplot}
        \caption{Projection in $\mathbb{R}^3$}
        \label{fig:3dplot}
    \end{subfigure}
\quad
 %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter}
        \caption{$Y_kV$}
        \label{fig:2dscatter}
    \end{subfigure}
\quad 
    \begin{subfigure}[b]{0.25\textwidth}
        \includegraphics[width=\textwidth]{Figs/2d-scatter-orth}
        \caption{$Y_kV_{\perp}$}
        \label{fig:2dscatterorth}
    \end{subfigure}
    \caption{ Two groups of four-dimensional data (red and blue)
      projected into different subspaces.  a) To visualize $Y_k$ we
      can project the data into $\mathbb{R}^3$.  In this illustration, the
      distributional differences between the groups are confined to a
      two-dimensional shared subspace ($VV^T$, grey plane).  b) The
      data projected onto the two-dimensional shared subspace,
      $Y_kV$, have covariances $\Psi_k$ that differ between
      groups. c) The orthogonal projection, $Y_kV_{\perp}$
      has isotropic covariance, $\sigma_k^2I$, for all groups.  }
\label{fig:shared}
\end{figure}
% For multi-group covariance estimation, we assume that we have $K$
% matrices, $Y_1, ... Y_K$, where $Y_k$ consists of $n_k$ independent
% rows of mean-zero normal data of dimension $p$, typically with
% $n_k \ll p$.

@@

\begin{align}
\nonumber p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}S_k/2)  \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\Psi_kV^T +
  I))^{-1}S_k/2) \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\Psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]S_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]S_k/2) ,
\end{align}
%
\noindent where we define $M_k = \sigma_k^2(\Psi_k + I)$.  The log-likelihood in
$V$ (up to an additive constant) is
%
\begin{align}
\nonumber l(V) &= \sum_k \tr\left(-VM_k^{-1}V^T +
       VV^T/\sigma^2_k\right)S_k/2)\\
&= \frac{1}{2}\sum_k \tr\left(-M_k^{-1}V^T S_kV\right) + \text{tr}\left(VV^T S_k\right)/\sigma_k^2.
\end{align}

% \begin{align}
% p(V|...) &\propto \etr(\sum_{k=1}^K[O_k\Omega_kO^T_k]V^TS_k/(2\sigma^2_k)]V)\\
% \end{align}

We maximize the marginal likelihood of $V$ with an EM algorithm, where
$M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ are considered the
``missing'' parameters.  We assume independent Jeffreys'
prior distributions for both $\sigma_k^2$ and $M_k$.  Jeffreys' prior
for these quantities corresponds to
$p(\sigma_k^2) \propto 1/\sigma_k^2$ and
$p(M_k) \propto |M_k|^{-(s+1)/2}$.  From the likelihood it can easily
be shown that the conditional posterior for $M_k$ is
%
$$p(M_k | V) \propto |M_k|^{(n_k + s + 1)/2}\etr(-(M_k^{-1}V^TS_kV)/2) $$
%
\noindent which is an inverse-Wishart($V^TS_kV$, $n_k$) distribution.  The
conditional posterior distribution of $\sigma_k^2$ is simply
%
$$p\left(\sigma^2 | V\right) \propto (\sigma_k^2)^{-n_k(p-s)/2-1}\etr\left(- (I-VV^T)S_k/[2\sigma_lk^2]\right)  $$
%
\noindent which is an inverse-gamma($n_k(p-s)/2$,
$\text{tr}[(I-VV^T)S_k]/2$) distribution.  Based on these
results, we complete the following steps for each iteration $t$ of
the EM algorithm until convergence:

\begin{enumerate}
\item For each $k$, compute relevant conditional expectations:
\begin{enumerate}
\item $E[M_k^{-1} | V_{(t-1)}] = n_k(V_{(t-1)}^T S_kV_{(t-1)})^{-1}$
\item $E[\frac{1}{\sigma_k^2}|V_{(t-1)}] = \frac{n_k(p-s)}{\text{tr}[(I-V_{(t-1)}V_{(t-1)}^T)S_k]}$
\end{enumerate}
\item Compute $V_{t}$ = $\underset{V}{\text{arg } \text{max}}  \sum_k \text{tr}\left(-VE[M^{-1}|V_{(t-1)}]^T +
       E[\frac{1}{2\sigma_k^2}|V_{(t-1)}]VV^T\right)S_k/2)$ 
\end{enumerate}

@@

\begin{figure}[t]
    \centering
%    \begin{subfigure}[b]{0.45\textwidth}
    \includegraphics[width=0.5\textwidth]{Figs/posteriorRegions-chull}
        \caption{95\% posterior regions}
        \label{fig:simPosterior}
%    \end{subfigure}
    ~ %add desired spacing between images, e. g. ~, \quad, \qquad, \hfill etc. 
      %(or a blank line to force the subfigure onto a new line)
    % \begin{subfigure}[b]{0.4\textwidth}
    %     \includegraphics[width=\textwidth]{Figs/simRatio-05-19}
    %     \caption{Ratio}
    %     \label{fig:simRatio}
    % \end{subfigure}
    \caption{a) 95\% posterior regions for the log of the ratio of
      eigenvalues, $\text{log}(\frac{\lambda_1}{\lambda_2})$, of $\Psi_k$ and the
      orientation of the principal axis on the space spanned by $\hat{V}$
      cover the truth in this simulation.  Dots correspond to true data generating
      parameter values on $\hat{V}^T\Sigma_k\hat{V}$ .  Since $V$
      is only identifiable up to rotation, for this figure we find the Procrustes
      rotation that maximizes the similarity of $\hat{V}$ to the true
      data generating basis. True eigenvalue ratios were 10 (red and
      black), 3 (green and blue) and 1 (light blue).  True
      orientations were $\pi/4$ (black), $-\pi/4$ (red and green) and
      0 (blue and cyan). % b) Goodness of fit,
      % $\gamma(Y_k: \hat{V}, \hat{\sigma}^2_k$ (Equation \ref{eqn:ratio}).
      % As expected, for each group, the inferred subspace explains
      % practically all of the variance from the first two eigenvectors
      % of $\Sigma_k$. 
    }
\end{figure}

@@

\begin{figure}[t]
    \centering
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{Figs/LossVsDimension}
        \caption{Stein's risk vs $\hat{s}$}
        \label{fig:sdimension}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{Figs/simRatio-s5}
        \caption{$\hat{s}$ = 5}
        \label{fig:ratio-s5}
    \end{subfigure}
    \begin{subfigure}[b]{0.3\textwidth}
        \includegraphics[width=\textwidth]{Figs/simRatio-s20}
        \caption{$\hat{s}$ = 20}
        \label{fig:ratio-s20}
    \end{subfigure}
    \caption{a) Stein's risk as a function of the
      shared subspace dimension.  Data from ten groups, with $U_k$
      generated uniformly on the Stiefel manifold
      $\mathcal{V}_{200, 2}$.  As $\hat{s} \rightarrow p$, the risk converges to the
      risk from independently estimated spiked covariance matrices
      (dashed blue line).  The data also fit a shared subspace model
      with $s=rK$.  If $VV^T = \text{span}(U_1, ..., U_k)$ were known
      exactly, shared subspace estimation yields lower risk than
      independent covariance estimation (dashed red line).  b) For a
      single simulated dataset, the goodness of fit statistic,
      $\gamma(Y_k: \hat{V}, \hat{\sigma_k}^2)$, when the assumed
      shared subspace is dimension $\hat{s} = 5$.  c).  For the same
      dataset, goodness of fit when the assumed shared subspace is
      dimension $\hat{s} = 20$.  We can capture nearly all of the
      variability in each of the 10 groups using an $\hat{s}=rK=20$
      dimensional shared subspace. }
\label{fig:dimensionPlots}
\end{figure}

@@

\begin{table}
\begin{center}
  \begin{tabular}{ l  l | c | c | c |}
    \multicolumn{2}{c}{} & \multicolumn{3}{c}{\textbf{Inferential Model}} \\
  \multicolumn{2}{c|}{}  & Adaptive & $\hat{\Sigma}_k=\hat{\Sigma}$
                                                           & $\hat{s} = p$ \\  \cline{2-5}
    \multirow{3}{*}{\rotatebox[origin=c]{90}{\textbf{Data Model}}} 
& $s=r=2$ & 0.8 (0.7, 0.9) & 2.1 (1.7, 2.6) & 3.0 (2.9, 3.2) \\ \cline{2-5}
   &   $s=r=2$, $\Sigma_k = \Sigma$ & 0.8 (0.7, 0.9) & 0.7 (0.6, 0.8) & 3.0 (2.9, 3.2)\\ \cline{2-5}
   &  $s=p=200$ & 7.1 (6.2, 8.0) & 138.2 (119, 153) & 3.0 (2.9, 3.2) \\ \cline{2-5}
  \end{tabular}
  \caption[Table caption text]{Stein's risk (and 95\% loss intervals)
    for different inferential models and data generating models with
    varying degrees of between-group covariance similarity.  For each
    of $K=10$ groups, we simulate data from three different types of
    shared subspace models.  For each of these models, $p=200$, $r=2$,
    $\sigma_k^2=1$ and $n_k=50$.  We also fit the data using three
    different shared subspace models: a model in which $s$, $r$ and
    $VV^T$ are all estimated from the data (``adaptive''), a spiked
    covariance model in which the covariance matrices from each group
    are assumed to be identical ($\hat{\Sigma}_k=\hat{\Sigma}$) and a
    model in which we assume the data do \emph{not} share a lower
    dimensional subspace across groups (i.e. $\hat{s} = p$). The estimators
    which most closely match the data generating model have the lowest
    risk (diagonal) but the adaptive estimator performs well relative
    to the alternative misspecified model.}
\label{table:groupLoss}
\end{center}
\end{table}

@@

>>>>

\documentclass[12pt]{article}
\pdfoutput=1
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{latexsym}
\usepackage{textcomp}
\usepackage{longtable}
\usepackage{multirow, booktabs}
\usepackage{natbib}
\usepackage{url}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{chemarrow}
\usepackage{subfigure}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{tikz}
\usepackage[margin=.75in]{geometry}

@@

\title{Shared Subspace Models for Multi-Group Covariance Estimation
%
  \protect\thanks{Alexander M. Franks is a Moore/Sloan Data Science
    and WRF Innovation in Data Science Postdoctoral Fellow
    (\href{mailto:amfranks@uw.edu}{amfranks@uw.edu}).  Peter D. Hoff
    is a Professor in the Department of Statistical Science at Duke University
    (\href{mailto:peter.hoff@duke.edu}{peter.hoff@duke.edu}).
%
This work was partially supported 
 by the Washington Research Foundation Fund for Innovation in
 Data-Intensive Discovery, the Moore/Sloan Data Science Environments
 Project at the University of Washington, NSF grant DMS-1505136, and NIH grant R03-CA211160-01.
%
 The authors are grateful to Dr.\ Daniel Promislow (Department of
 Pathology, University of Washington), Dr.\ Jessica Hoffman
 (University of Alabama at Birmingham) and Dr. Julie Josse (INRIA)
 for sharing data and ideas which contributed to framing of this
 paper.}}  \author{Alexander Franks and Peter Hoff} \date{\today}
\maketitle

\begin{abstract}

@@

\begin{figure}[t]
    \centering
    \subfigure[Projection in $\mathbb{R}^3$]{
    \label{fig:3dplot}
    \includegraphics[width=0.25\textwidth]{Figs/3dplot}}
\quad
    \subfigure[$Y_kV$]{
    \label{fig:2dscatter}
    \includegraphics[width=0.25\textwidth]{Figs/2d-scatter}}
\quad 
\subfigure[$Y_kV_{\perp}$]{
\label{fig:2dscatterorth}
\includegraphics[width=0.25\textwidth]{Figs/2d-scatter-orth}}
\caption{ Two groups of four-dimensional data (red and blue)
      projected into different subspaces.  a) To visualize $Y_k$ we
      can project the data into $\mathbb{R}^3$.  In this illustration, the
      distributional differences between the groups are confined to a
      two-dimensional shared subspace ($VV^T$, grey plane).  b) The
      data projected onto the two-dimensional shared subspace,
      $Y_kV$, have covariances $\Psi_k$ that differ between
      groups. c) The orthogonal projection, $Y_kV_{\perp}$
      has isotropic covariance, $\sigma_k^2I$, for all groups.  }
\label{fig:shared}
\end{figure}
% For multi-group covariance estimation, we assume that we have $K$
% matrices, $Y_1, ... Y_K$, where $Y_k$ consists of $n_k$ independent
% rows of mean-zero normal data of dimension $p$, typically with
% $n_k \ll p$.

@@

\begin{align}
\nonumber p(S_1, ... S_k | \Sigma_k,n_k) &\propto \prod_{k=1}^K |\Sigma_k|^{-n_k/2}\etr(-\Sigma_k^{-1}S_k/2)  \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-(\sigma_k^2(V\Psi_kV^T +
  I))^{-1}S_k/2) \\
\nonumber &\propto \prod_{k=1}^K  |\Sigma_k|^{-n_k/2}\etr(-\left[V(\Psi_k +
  I)^{-1}/\sigma_k^2 V^T + (I-VV^T)/\sigma^2_k\right]S_k/2)
  \\
&\propto \prod_{k=1}^K  (\sigma_k^2)^{-n_k(p-s)/2}|M_k|^{-n_k/2}\etr(-\left[VM_k^{-1}V^T + \frac{1}{\sigma^2_k} (I-VV^T)\right]S_k/2) ,
\end{align}
%
\noindent where we define $M_k = \sigma_k^2(\Psi_k + I)$.  The log-likelihood in
$V$ (up to an additive constant) is
%
\begin{align}
\nonumber l(V) &= \sum_k \tr\left(-VM_k^{-1}V^T +
       VV^T/\sigma^2_k\right)S_k/2)\\
%\nonumber &= \frac{1}{2}\sum_k \tr\left(-M_k^{-1}V^T S_kV\right) +
%  \text{tr}\left(VV^T S_k\right)/\sigma_k^2\\
&=\frac{1}{2}\sum_k \tr\left((\frac{1}{\sigma_k^2}I-M_k^{-1})V^T
  S_kV\right).
\label{eqn:likV}
\end{align}
%
We maximize the marginal likelihood of $V$ with an EM algorithm, where
$M_k^{-1}$ and $\frac{1}{\sigma_k^2}$ are considered the
``missing'' parameters.  We assume independent Jeffreys'
prior distributions for both $\sigma_k^2$ and $M_k$.  Jeffreys' prior
for these quantities corresponds to
$p(\sigma_k^2) \propto 1/\sigma_k^2$ and
$p(M_k) \propto |M_k|^{-(s+1)/2}$.  From the likelihood it can easily
be shown that the conditional posterior for $M_k$ is
%
$$p(M_k | V) \propto |M_k|^{(n_k + s + 1)/2}\etr(-(M_k^{-1}V^TS_kV)/2) $$
%
\noindent which is an inverse-Wishart($V^TS_kV$, $n_k$) distribution.  The
conditional posterior distribution of $\sigma_k^2$ is simply
%
$$p\left(\sigma^2 | V\right) \propto (\sigma_k^2)^{-n_k(p-s)/2-1}\etr\left(- (I-VV^T)S_k/[2\sigma_k^2]\right)  $$
%
\noindent which is an inverse-gamma($n_k(p-s)/2$,
$\text{tr}[(I-VV^T)S_k]/2$) distribution.  Based on these
results, we complete the following steps for each iteration $t$ of
the EM algorithm until convergence:

\begin{enumerate}
\item For each $k$, compute relevant conditional expectations:
\begin{enumerate}
\item $E[M_k^{-1} | V_{(t-1)}] = n_k(V_{(t-1)}^T S_kV_{(t-1)})^{-1}$
\item $E[\frac{1}{\sigma_k^2}|V_{(t-1)}] = \frac{n_k(p-s)}{\text{tr}[(I-V_{(t-1)}V_{(t-1)}^T)S_k]}$
\end{enumerate}
\item Compute $V_{t}$ = $\underset{V}{\text{arg } \text{max}}  \sum_k \text{tr}\left(-VE[M^{-1}|V_{(t-1)}]V^T +
       E[\frac{1}{2\sigma_k^2}|V_{(t-1)}]VV^T\right)S_k/2)$ 
\end{enumerate}

@@

\begin{figure}[t]
    \centering
    \includegraphics[width=0.5\textwidth]{Figs/posteriorRegions-chull}
        \label{fig:simPosterior}
    \caption{a) 95\% posterior regions for the log of the ratio of
      eigenvalues, $\text{log}(\frac{\lambda_1}{\lambda_2})$, of $\Psi_k$ and the
      orientation of the principal axis on the space spanned by $\hat{V}$
      cover the truth in this simulation.  Dots correspond to true data generating
      parameter values on $\hat{V}^T\Sigma_k\hat{V}$ .  Since $V$
      is only identifiable up to rotation, for this figure we find the Procrustes
      rotation that maximizes the similarity of $\hat{V}$ to the true
      data generating basis. True eigenvalue ratios were 10 (red and
      black), 3 (green and blue) and 1 (light blue).  True
      orientations were $\pi/4$ (black), $-\pi/4$ (red and green) and
      0 (blue and cyan). % b) Goodness of fit,
      % $\gamma(Y_k: \hat{V}, \hat{\sigma}^2_k$ (Equation \ref{eqn:ratio}).
      % As expected, for each group, the inferred subspace explains
      % practically all of the variance from the first two eigenvectors
      % of $\Sigma_k$. 
    }
\end{figure}

@@

\begin{figure}[t]
    \centering
    \subfigure[Stein's risk vs $\hat{s}$]{
      \label{fig:sdimension}
      \includegraphics[width=0.3\textwidth]{Figs/LossVsDimension}}
    \subfigure[$\hat{s}$ = 5]{
      \label{fig:ratio-s5}
      \includegraphics[width=0.3\textwidth]{Figs/simRatio-s5}}
    \subfigure[$\hat{s}$ = 20]{
        \label{fig:ratio-s20}
        \includegraphics[width=0.3\textwidth]{Figs/simRatio-s20}}
      \caption{a) Stein's risk as a function of the
      shared subspace dimension.  Data from ten groups, with $U_k$
      generated uniformly on the Stiefel manifold
      $\mathcal{V}_{200, 2}$.  As $\hat{s} \rightarrow p$, the risk converges to the
      risk from independently estimated spiked covariance matrices
      (dashed blue line).  The data also fit a shared subspace model
      with $s=rK$.  If $VV^T = \text{span}(U_1, ..., U_k)$ were known
      exactly, shared subspace estimation yields lower risk than
      independent covariance estimation (dashed red line).  b) For a
      single simulated dataset, the goodness of fit statistic,
      $\gamma(Y_k: \hat{V}, \hat{\sigma_k}^2)$, when the assumed
      shared subspace is dimension $\hat{s} = 5$.  c).  For the same
      dataset, goodness of fit when the assumed shared subspace is
      dimension $\hat{s} = 20$.  We can capture nearly all of the
      variability in each of the 10 groups using an $\hat{s}=rK=20$
      dimensional shared subspace. }
\label{fig:dimensionPlots}
\end{figure}

@@

\begin{table}
  \caption[abcd]{Stein's risk (and 95\% loss intervals)
    for different inferential models and data generating models with
    varying degrees of between-group covariance similarity.  For each of $K=10$ groups, we simulate data from three different types of shared subspace models.  For each of these models, $p=200$, $r=2$,
    $\sigma_k^2=1$ and $n_k=50$.  We also fit the data using three
    different shared subspace models: a model in which $s$, $r$ and
    $VV^T$ are all estimated from the data (``adaptive''), a spiked
    covariance model in which the covariance matrices from each group
    are assumed to be identical ($\hat{\Sigma}_k=\hat{\Sigma}$) and a
    model in which we assume the data do \emph{not} share a lower
    dimensional subspace across groups (i.e. $\hat{s} = p$). The estimators
    which most closely match the data generating model have the lowest
    risk (diagonal) but the adaptive estimator performs well relative
    to the alternative misspecified model. \label{table:groupLoss}}
\centering
 \begin{tabular}{ l  l | c | c | c |}
    \multicolumn{2}{c}{} & \multicolumn{3}{c}{\textbf{Inferential Model}} \\
  \multicolumn{2}{c|}{}  & Adaptive & $\hat{\Sigma}_k=\hat{\Sigma}$
                                                           & $\hat{s} = p$ \\  \cline{2-5}
    \multirow{3}{*}{\rotatebox[origin=c]{90}{\textbf{Data Model}}} 
& $s=r=2$ & 0.8 (0.7, 0.9) & 2.1 (1.7, 2.6) & 3.0 (2.9, 3.2) \\ %\cline{2-5}
   &   $s=r=2$, $\Sigma_k = \Sigma$ & 0.8 (0.7, 0.9) & 0.7 (0.6, 0.8) & 3.0 (2.9, 3.2)\\ %\cline{2-5}
   &  $s=p=200$ & 7.1 (6.2, 8.0) & 138.2 (119, 153) & 3.0 (2.9, 3.2) \\ %\cline{2-5}
  \end{tabular}
\end{table}

@@

\section{Reduction of Asymptotic Bias Via Pooling}
\label{sec:asymp}
Recently, there has been an interest in the asymptotic behavior of
PCA-based covariance estimators in the setting in which
$p, n \to \infty$ with $p/n=\gamma$ fixed.  Specifically, in the
spiked covariance model it is known that when $p$ and $n$ are both
large, the leading eigenvalues of the sample covariance matrix are
positively biased and the empirical eigenvectors form a non-zero angle
with the true eigenvectors \citep{Baik2006, Paul2007}.  Although this
fact also implies that the shared subspace estimators are biased, a
major advantage of shared subspace inference over independent
estimation of multiple covariance matrices is that we reduce the
asymptotic bias by pooling information across groups.  The bias
reduction is especially large when there is significant heterogeneity
in the first $s$ eigenvectors of the projected covariance matrices.

Throughout section we assume $K$ groups of data each with $n_k = n$
observations per group and $s$ a fixed constant.  First, note that if
$\hat{V}\hat{V}^T$ corresponds to the true shared subspace, then
estimates $\hat{\psi}_k$ derived using the methods presented in
Section \ref{sec:bayes} will consistently estimate $\psi_k$ as
$n \to \infty$ regardless of whether $p$ increases as well because
$Y_kV$ has a fixed number of columns.  For this reason, we focus
explicitly on the accuracy of $\hat{V}\hat{V}^T$ (derived using the
maximum marginal likelihood algorithm presented in Section
\ref{sec:em}) as a function of the number of groups $K$ when both $p$
and $n$ are of the same order of magnitude and much larger than $s$.
As an accuracy metric, we study the behavior of
$\tr(\hat{V}\hat{V}^TVV^T)/s$.  This metric is bounded by zero and one
and achieves a maximum of one if and only if $\hat{V}\hat{V}^T$
corresponds to the true shared subspace.

To start, we consider the simple case in which the data matrices $Y_k$
are identically distributed with covariance $\Sigma_k=\Sigma$ of the
shared-subspace form given in Equation \ref{eqn:sspsi} and
$\sigma_k^2 = 1$.  Without loss of generality we can let
$\psi_k = \psi$ be a diagonal matrix (e.g assume the columns of $V$
align with the eigenvectors of $\Sigma$).  In this case, for a single
group the complete data likelihood of $V$ (Equation \ref{eqn:likV})
can be rewritten as
%
\begin{align*}
\ell(V) &=\frac{1}{2}\tr\left((\frac{1}{\sigma^2}I-M^{-1})V^T
  S_kV\right)\\
&=\frac{1}{2}\tr\left(DV^TS_kV\right).
\end{align*}
%
Since $\psi$ is diagonal and $\sigma^2=1$, 
$M = \sigma^2(\psi + I)$ is diagonal and thus
$D = (\frac{1}{\sigma^2}I-M^{-1})$ is also diagonal with entries
$0 < d_i < 1$ of decreasing magnitude.  Then, the solution to
$$\hat{V}^{(k)} = \underset{\widetilde{V} \in \mathcal{V}_{p,
    s}}{\text{argmax }} \tr\left(D\widetilde{V}^T
  S_k\widetilde{V}\right).$$
has $\hat{V}^{(k)}$ equal to the first $s$ eigenvectors of $S_k$.  This is
maximized when the columns of $V$ match the first empirical eigenvectors of
$S_k$ and has a maximum of $\sum_{i=1}^r d_i\ell_i$ where $\ell_i$ is the
$i$th eigenvalue of $S_k$.  As shown in \citet{Paul2007}, as long as
$\lambda_i > 1 + \sqrt{\gamma}$ where $\lambda_i$ is the $i$th
eigenvalue of $\Sigma_k$, the asymptotic inner product between
the $i$th sample eigenvector and the $i$th population eigenvector
approaches a limit that is almost surely less than one
$$|\langle\hat{V}^{(k)}_i, V_i\rangle| \to \sqrt{\left(1-\frac{\gamma}{(\lambda_i - 1)^2}\right) /\left(1 +
    \frac{\gamma}{(\lambda_i - 1)}\right)} $$
%
% Note that in the likelihood specified in Equation \ref{eqn:likV} , the
% sample covariance matrices, $S_k/n_k$, are sufficient statistics.
% Since the sample covariance matrices are biased for the true
% eigenvectors of $\Sigma_k$ when $p$ and $n$ increase at the same rate
% then $\tr(\hat{V}\hat{V}^TVV^T)/s < 1$ almost surely \citep{}.

\noindent As such, for each independent estimate
$\hat{V}^{(k)}\hat{V}^{(k)^T}$, we can approximate the shared subspace
accuracy as
\begin{align}
\nonumber \tr(\hat{V}^{(k)}\hat{V}^{(k)^T}VV^T)/s &= \frac{1}{s}\sum_{i=1}^s |\langle\hat{V}^{(k)}_i, V_i\rangle|^2\\
&\approx \frac{1}{s}\sum_{i=1}^s  \left(1-\frac{\gamma}{(\lambda_i - 1)^2}\right) /\left(1 +
    \frac{\gamma}{(\lambda_i - 1)}\right).
\label{eqn:lbound}
\end{align}

Naturally, when the groups are identically distributed, $VV^T$ should
be estimated using the pooled data.  In this simple case, it is easy
to analytically express the improvement in subspace accuracy from
pooling data across groups.  The maximum likelihood estimate for
$VV^T$ for the pooled data can be expressed as
\begin{align*}
\hat{V} &= \underset{\widetilde{V} \in \mathcal{V}_{p,
    s}}{\text{argmax }} \tr\left(D\widetilde{V}^T
  (\sum_{k=1}^K S_k)\widetilde{V}\right)
\end{align*}
%
\noindent where $\sum_{k=1}^K S_k \sim \text{Wish}(\Sigma, Kn)$.
Thus, in this example we can express the accuracy of
$\hat{V}\hat{V}^T$ for the pooled estimator as:
%
\begin{align}
\nonumber \tr(\hat{V}\hat{V}^TVV^T)/s &= \sum_{i=1}^s |\langle\hat{V}_i, V_i\rangle|^2\\
&\approx\frac{1}{s}\sum_{i=1}^s  \left(1-\frac{\gamma}{K(\lambda_i - 1)^2}\right) /\left(1 +
    \frac{\gamma}{K(\lambda_i - 1)}\right).
\label{eqn:bound}
\end{align}

Here, the accuracy of the estimate depends on $\gamma$, $K$ and the
magnitude of the eigenvalues, with the bias decreasing as the number
of groups $K$ increases.  Empirically, in the fully general setting in
which $\psi_k$ varies across groups we get bias reductions that scale
similarly in $K$.  This is supported by the results in Table
\ref{table:groupLoss} which show that when estimating groups of spiked
covariance matrices using the shared subspace estimator there is a
significant reduction in Stein's risk. Moreover, Equation
\ref{eqn:bound} provides a useful benchmark for understanding the
bias of shared subspace estimates.

\begin{figure}[t]
    \centering
    \includegraphics[width=0.7\textwidth]{Figs/asymptotics}
    \caption{Subspace accuracy $\tr(\hat{V}\hat{V}^TVV^T)/s$ (solid)
      and the asymptotics-based benchmark (dashed)
      % $\frac{1}{s}\sum_{i=1}^s \left(1-\frac{\gamma}{K(\lambda_i -
      %     1)^2}\right) /\left(1 + \frac{\gamma}{K(\lambda_i -
      %     1)}\right)$
      as a function of $K$.  When
      $\lambda_1=\lambda_2$ (green), the assumptions used to derive
      the benchmark (identically distributed groups) are met and thus the
      subspace accuracy matches the benchmark.  However, when the
      ratio $\lambda_1/\lambda_2$ is large, the subspace accuracy
      metric can far exceed this benchmark if there is significant
      variation in the eigenvectors of $\psi_k$ across groups.  Small
      increases in accuracy over the benchmark are seen for moderately
      anisotropic data (red) and large increases for highly
      anisotropic data (blue).  }
      \label{fig:asymptotics}
\end{figure}

In Figure \ref{fig:asymptotics} we depict the subspace accuracy metric
$\tr(\hat{V}\hat{V}^TVV^T)/s$ and benchmark
$\frac{1}{s}\sum_{i=1}^s \left(1-\frac{\gamma}{K(\lambda_i -
    1)^2}\right) /\left(1 + \frac{\gamma}{K(\lambda_i - 1)}\right)$
for simulated multi-group data generated under the shared subspace
model with $s=2$, $n=50$, $p=200$ and three different sets of eigenvalues.  For each
covariance matrix, the eigenvectors of $\psi_k$ were sampled uniformly
from Stiefel manifold $\mathcal{V}_{2,2}$.  When $\psi_k$ is isotropic
(green) the subspace similarity metric closely matches the benchmark
since the assumptions used to derive this asymptotic result are met.
However, when the eigenvectors of $\psi_k$ vary significantly across
groups and $\lambda_1 \gg \lambda_2$, the subspace accuracy can far
exceed this benchmark (blue).  Intuitively, when the first
eigenvectors of two different groups are nearly orthogonal, each group
provides a lot of information about orthogonal directions on $VV^T$
and so the gains in accuracy exceed those that you would get by
estimating the subspace from a single group with $K$ times the sample
size.  In general the accuracy of shared subspace estimates depends on
the variation in the eigenvectors of $\psi_k$ across groups as well as
the magnitude of the eigenvalues and matrix dimensions $p$ and $n_k$.
Although the shared subspace estimator improves on the accuracy of
individually estimated covariance matrices, estimates can still be
biased when $\gamma$ is very large or the eigenvalues of $\Sigma_k$
are very small for all $k$.  In practice, one should estimate the
approximate magnitude of the bias using the inferred eigenvalues of $\Sigma_k$.  When
these inferred eigenvalues are significantly larger than
$\hat{\sigma}_k^2(1+\sqrt{\gamma/K})$ the bias will likely be small.

% \begin{align*}
% \hat{V} &= \underset{\widetilde{V} \in \mathcal{V}_{p,
%     s}}{\text{argmax }} \tr\left((I-M_k^{-1})^{1/2}\widetilde{V}^T
%   (\sum_{k=1}^K S_k)\widetilde{V}(I-M_k^{-1})^{1/2}\right)
% \end{align*}
% Where $(I-M_k^{-1})^{1/2}$

% \begin{align*}
% Y &= ZV\psi_k^{1/2}V^T+E\\
% Y^TY &= V\psi_k^{1/2}V^TZ^TZV\psi_k^{1/2}V^T + E^TE + \text{cross terms}
% \end{align*}

% \noindent where $Z$ and $E$ are $n$ by $p$ matrices of iid stanard
% normals.

% $$ \tr\left((I-M_k^{-1})\widetilde{V}^TVM_k^{1/2}V^TZ^TZVM_k^{1/2}V^T\widetilde{V}\right)$$
%% \subsection{Consistency of $\Psi_k$ and $\sigma_k^2$ given $V$}
